The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-07-31 analysis/paper/paper.Rmd
Reordered discussion

Bipolar technology clearly had an important role on the first modern humans in Western Iberia. As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015]. For the first modern humans, as well as their descendants, this technology would have played a major role in resource exploitation and diversification. In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease-rendering techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the constant correlation between red deer bone fragments with the amount of bone marrow and grease contained in them. Other functional indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks that were used as anvils.

In Table 6 we present current data based on previous work on bipolar technology in Westernmost Iberia for the Upper Paleolithic. While the number of artifacts is not very high, bipolar technology is present in a significant number of sites. Even with the considerable number of sites represented, it is clear the discrepancy in counts when comparing the presence of scaled pieces and bipolar cores, which by itself could be indicative of some sort of functional pattern. It is clear, however, that scaled pieces show on average, a higher representation in Vale Boi’s assemblages than on most other sites. Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.  While currently there are no data on the Typological Index (TI) of the Slope’s scaled pieces, since the rest of the assemblage has not yet been studied, it is fairly safe to assume that the TI would have a particularly high value. 
Bipolar technology is quite common in Western Iberian Upper Paleolithic sites and interpretations made are solely based on a single reference with no statistical or analytical evidence. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different, although these interpretations should be in the future backed by some sort of analytical evidence. As our results show, the low presence of bipolar cores is quite evident in these sites, while scaled pieces are more often present. It can also be argued that the number of pieces in the overall picture is also low, although presenting an accurate picture for each site since most of these sites have been extensively studied.

@@

Several researchers argued that scaled pieces were used as cores for the extraction of chips or small bladelets [e.g. @carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997]. Here, we argue that the concept itself is flawed, especially when applied to Vale Boi. With the single exception of the early Gravettian, the site’s technology shows no evidence of manufacture or use of tools with such small dimensions (5-8 mm). The most common interpretation is that these chips or bladelets would be inserted in composite throwing tools. If we look at the average length of Vale Boi’s chips removed from scaled pieces we have an average of 7 mm. The functionality of chips of this size is quite arguable, even when inserted into a composite tool, and especially the impact that these would have in a perforation or extension of bleeding scenario opposed to using regular bladelets. Also, there is no evidence of bone or antler tools with perforation marks for the insertion of chips, although this kind of weapon could have been exclusively made with perishable materials such as wood. Regarding the technology, as previously mentioned, in Vale Boi every phase of the chaîne operatoire is present, which means the knapping was made in situ, and so the number of chips found is quite high. Considering that most knapping activities originate chips, the need of a specific tool for the sole purpose of extracting them does not seem viable. By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities (e.g. working wood or antler with the goal of producing utensils) seems to be the most reasonable interpretation.

Our results also suggest that higher degrees of representation of scaled pieces in sites can be originated by factors such as site functionality but also by cultural patterns. Regarding site functionality, if we look at the sites with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites. One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c. 17,12% of the total assemblage. This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site. However, these data can be misleading, as Casal do Felipe (TI=5,91), Terra do Manuel 1940-42 (TI=1,03), Terra do Manuel 1988-89 (TI=2,13) and Vale Comprido – Encosta (TI=0,9) are also open-air multifunctional sites. These sites show the same functionality yet, smaller representations of scaled pieces.

>>>>

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia and elsewhere. As mentioned by Shot and Tostevin [-@shott_diversity_2015], a rise in bipolar technology frequency may be related to human mobility patterns and even evolutionary trends [@duke_weaving_2015]. For the first modern humans, as well as their descendants, this technology would have played a major role in the intensive and diverse exploitation of resources. In Vale Boi, bipolar technology shows up in the archaeological record connected to evidence related to the intensification and diversification of resource exploration such as the above-mentioned grease obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents. Other  indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils.

@@

Several researchers argued that scaled pieces were used as cores for the extraction of chips or small bladelets [e.g. @carvalho_talhe_1998; @de_la_pena_sobre_2011; @leblanc_wedges_1992; @shott_bipolar_1989; @shott_bipolar_1999; @zilhao_o_1997]. Based on our data we find that this concept is somehow flawed, especially when applied to Vale Boi. First, with the single exception of the early Gravettian (during which very small backed bladelets are present - see Marreiros et al [-@marreiros_lithic_2018]), the site’s technology shows no evidence of manufacture or use of tools with such small dimensions as the ones recorded from the scars of scaled pieces (5-8 mm). Second, although the most common interpretation is that these chips/bladelets would be inserted in composite throwing tools, the functionality of chips of ~7mm is quite arguable, especially regarding the impact that these would have in a perforation or extension of a bleeding scenario, as opposed to using regular bladelets. Third, although Vale Boi has a very rich assemblage of organic tools [@evora_osseous_2013], there is no evidence of bone or antler tools with perforation marks for the insertion of stone implements. Finally, as previously mentioned, Vale Boi stone tool technological analyses revealed that every phase of the chaîne operatoire is present, which means that knapping was mostly occurring at the site, which is attested by the high number of chips found. Considering that most knapping activities originate chipage, the need of a specific tool for the sole purpose of extracting them does not seem viable to us. By putting all these factors in context, the hypothesis that these pieces were used for carcass processing, bone marrow extracting [@manne_intensive_2012] and other similar activities, such as working hard, organic materials with the goal of producing tools seems, thus far, to be the most reasonable interpretation.

In Table 6 we present current data on Upper Paleolithic bipolar technology in Portugal. Although the number of scaled pieces and bipolar cores is considerably low in most of the sites, a significant number of *loci* present some kind of bipolar technology evidence. Moreover, it is still clear a discrepancy when comparing the presence of scaled pieces versus bipolar cores, the latter being well under represented than the first. It is clear, also, that scaled pieces show on average a higher representation in Vale Boi’s assemblages than on most other sites. Further, of the three bipolar cores in all sites, 2 of these come from Vale Boi.  While currently there are no definitive data on the representation of scaled pieces within the retouched tool assemblage of the Slope area, it is fairly safe to assume that it would have a particularly high value.

Bipolar technology is quite common in Upper Paleolithic sites of Western Iberia, but current interpretations are based on empirical observations, rather than on more thoroughly analytical evidence. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different.

As our results show, the low presence of bipolar cores is quite evident in these sites, while scaled pieces are more often present. It can also be argued that the number of pieces in the overall picture is also low, although presenting an accurate picture for each site since most of these sites have been extensively studied.

Higher degrees of representation of scaled pieces in sites can be originated by factors such as site function, but also by cultural patterns. Regarding site functionality, if we look at the *loci* with higher counts of scaled pieces (Vale Boi’s Terrace and Slope and Fonte Santa), both are open-air residential sites. One example outside of this region is the Southwestern German Magdalenian site of Gönnersdorf, where according to Sano (2012) scaled pieces were the second largest assemblage from the site’s retouched utensils (257 from a total of 1501 utensils), corresponding to c. 17,12% of the total assemblage. This typological index is as high as both Vale Boi’s and Fonte Santa’s and like both of these sites, Gönnersdorf is an open-air residential site. However, these data can be misleading, as Casal do Felipe (TI=5,91), Terra do Manuel 1940-42 (TI=1,03), Terra do Manuel 1988-89 (TI=2,13) and Vale Comprido – Encosta (TI=0,9) are also open-air multifunctional sites. These sites show the same functionality yet, smaller representations of scaled pieces.
